\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{array}
\usepackage{float}
\title{Final Project: Naive Bayesian Classifier and Bagging}
\author{}
\date{}

\begin{document}
\maketitle

\section{Introduction}
This project evaluates a Naive Bayesian classifier (NBC) and bagging ensembles of NBC models on three UCI Machine Learning Repository datasets: Banknote Authentication, Glass Identification, and Image Segmentation. The goal is to compare the original NBC with bagging ensembles containing 10, 20, and 30 base models.

\section{Methodology}
All datasets were downloaded by the \texttt{ucimlrepo} package. The dataset identifiers were 267 for Banknote Authentication, 42 for Glass Identification, and 50 for Image Segmentation. Continuous attributes were discretized with equal-width discretization into ten bins. The discretizer was fit only on the training portion of each fold, then applied to the corresponding test portion.

The NBC model estimates class priors and feature conditional probabilities with Laplace estimation. Missing feature values are ignored when estimating conditional probabilities and are also skipped during prediction. For bagging, each base NBC was trained on a bootstrap sample of the training fold. Final predictions were produced by majority vote. Ties were broken deterministically by the sorted class labels.

Performance was measured by five-fold cross validation using only train and test splits. The fold assignment follows Chapter 4 method III. First, each instance is assigned a random number from the uniform interval [0,1]. Second, instances are sorted by the random number. Third, the sorted instances are read sequentially and assigned cyclically to folds 1 through 5. This makes the fold sizes approximately equal, and the size difference between any two folds is at most one. The random seed was fixed at 42 for reproducibility.

Four Chapter 5 comparisons were computed. For one data set, data-set aggregation pools correct predictions over all test folds and applies Theorem 5.1's two-proportion statistic
\[
z=\frac{\hat p_2-\hat p_1}{\sqrt{2\hat p(1-\hat p)/n}},
\qquad \hat p=\frac{\hat p_1+\hat p_2}{2}.
\]
Fold aggregation uses the matched differences $d_l=\hat p_{2l}-\hat p_{1l}$ from the same folds and Theorem 5.2's statistic $t=\bar d/\sqrt{s_d^2/k}$ with $k-1$ degrees of freedom. For multiple data sets, Theorem 5.4 averages each data set's pooled accuracy difference with equal data-set weight and estimates its normal variance from each data set's size. Theorems 5.5 and 5.6 average matched fold differences inside each data set and then across data sets; their $t$ statistic combines the within-data-set difference variances, and the degrees of freedom are rounded down according to Theorem 5.6. Positive differences in all tables mean that the comparison model is more accurate than the baseline. All data-set-level counts satisfy Chapter 5's large-sample condition.

\section{Experimental Results}
\begin{table}[H]
\centering
\small
\caption{Five-fold cross-validation accuracy.}
\label{tab:summary}
\begin{tabular}{llrr}
\toprule
Dataset & Model & Base models & Accuracy (\%) \\
\midrule
Banknote Authentication & NBC & 1 & 89.80 $\pm$ 3.21 \\
Banknote Authentication & Bagging NBC (10) & 10 & 89.65 $\pm$ 3.67 \\
Banknote Authentication & Bagging NBC (20) & 20 & 89.94 $\pm$ 3.67 \\
Banknote Authentication & Bagging NBC (30) & 30 & 90.09 $\pm$ 3.44 \\
Glass Identification & NBC & 1 & 60.29 $\pm$ 9.68 \\
Glass Identification & Bagging NBC (10) & 10 & 61.73 $\pm$ 12.46 \\
Glass Identification & Bagging NBC (20) & 20 & 62.60 $\pm$ 7.07 \\
Glass Identification & Bagging NBC (30) & 30 & 64.52 $\pm$ 11.34 \\
Image Segmentation & NBC & 1 & 80.00 $\pm$ 5.98 \\
Image Segmentation & Bagging NBC (10) & 10 & 79.05 $\pm$ 6.39 \\
Image Segmentation & Bagging NBC (20) & 20 & 80.00 $\pm$ 6.43 \\
Image Segmentation & Bagging NBC (30) & 30 & 80.00 $\pm$ 4.33 \\
\bottomrule
\end{tabular}
\end{table}

\section{Analysis}
The current run produced the same deterministic results whenever it was repeated with seed 42. For Banknote Authentication, NBC obtained 89.80\% mean fold accuracy and Bagging-30 obtained the highest value, 90.09\%, an increase of only 0.29 percentage points at the data-set aggregation level.

Glass Identification showed the largest numerical benefit. NBC obtained 60.29\% mean fold accuracy, compared with 61.73\%, 62.60\%, and 64.52\% for Bagging-10, Bagging-20, and Bagging-30. Thus, Bagging-30 improved the data-set-aggregated accuracy over NBC by 4.21 percentage points.

For Image Segmentation, NBC, Bagging-20, and Bagging-30 all produced 80.00\% mean accuracy, while Bagging-10 obtained 79.05\%. Bagging therefore did not improve the final accuracy on this data set under the selected preprocessing and model settings.

\section{Result Evaluation}
Table~\ref{tab:single-nbc} reports both single-data-set aggregation levels. None of the NBC-versus-bagging comparisons is significant at $\alpha=0.05$. The largest observed gain is NBC versus Bagging-30 on Glass Identification: the pooled data-set difference is 4.21 percentage points ($z=0.8980$, $p=0.3692$), and the matched fold difference is 4.23 points ($t=1.2097$, $df=4$, $p=0.2930$).

\begin{table}[H]
\centering
\scriptsize
\caption{Single-dataset NBC versus bagging comparisons. Positive differences favor bagging.}
\label{tab:single-nbc}
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lllrrrrrr}
\toprule
Dataset & Baseline & Comparison & Data diff. (\%) & $z$ & $p_z$ & Fold diff. (\%) & $t$ & $p_t$ \\
\midrule
Banknote Authentication & NBC & Bagging NBC (10) & -0.15 & -0.1257 & 0.8999 & -0.15 & -0.2812 & 0.7925 \\
Banknote Authentication & NBC & Bagging NBC (20) & 0.15 & 0.1265 & 0.8993 & 0.15 & 0.3866 & 0.7187 \\
Banknote Authentication & NBC & Bagging NBC (30) & 0.29 & 0.2539 & 0.7996 & 0.29 & 0.5577 & 0.6068 \\
Glass Identification & NBC & Bagging NBC (10) & 1.40 & 0.2973 & 0.7663 & 1.44 & 0.4002 & 0.7095 \\
Glass Identification & NBC & Bagging NBC (20) & 2.34 & 0.4966 & 0.6195 & 2.31 & 1.0463 & 0.3545 \\
Glass Identification & NBC & Bagging NBC (30) & 4.21 & 0.8980 & 0.3692 & 4.23 & 1.2097 & 0.2930 \\
Image Segmentation & NBC & Bagging NBC (10) & -0.95 & -0.2418 & 0.8089 & -0.95 & -0.7845 & 0.4766 \\
Image Segmentation & NBC & Bagging NBC (20) & 0.00 & 0.0000 & 1.0000 & 0.00 & 0.0000 & 1.0000 \\
Image Segmentation & NBC & Bagging NBC (30) & 0.00 & 0.0000 & 1.0000 & 0.00 & 0.0000 & 1.0000 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:single-bagging} compares ensemble sizes. The matched fold test finds one significant result: on Glass Identification, Bagging-30 exceeds Bagging-10 by 2.79 points ($t=3.2071$, $df=4$, $p=0.0327$). The data-set aggregation test for the same comparison is not significant ($z=0.6010$, $p=0.5479$). This difference illustrates that the aggregation level changes the sampling variance and therefore the inference.

\begin{table}[H]
\centering
\scriptsize
\caption{Single-dataset comparisons among bagging ensemble sizes.}
\label{tab:single-bagging}
\setlength{\tabcolsep}{3pt}
\begin{tabular}{lllrrrrrr}
\toprule
Dataset & Baseline & Comparison & Data diff. (\%) & $z$ & $p_z$ & Fold diff. (\%) & $t$ & $p_t$ \\
\midrule
Banknote Authentication & Bagging NBC (10) & Bagging NBC (20) & 0.29 & 0.2523 & 0.8008 & 0.29 & 1.0918 & 0.3363 \\
Banknote Authentication & Bagging NBC (10) & Bagging NBC (30) & 0.44 & 0.3796 & 0.7042 & 0.44 & 1.6352 & 0.1773 \\
Banknote Authentication & Bagging NBC (20) & Bagging NBC (30) & 0.15 & 0.1273 & 0.8987 & 0.15 & 0.5874 & 0.5885 \\
Glass Identification & Bagging NBC (10) & Bagging NBC (20) & 0.93 & 0.1993 & 0.8420 & 0.87 & 0.2200 & 0.8367 \\
Glass Identification & Bagging NBC (10) & Bagging NBC (30) & 2.80 & 0.6010 & 0.5479 & 2.79 & 3.2071 & 0.0327 \\
Glass Identification & Bagging NBC (20) & Bagging NBC (30) & 1.87 & 0.4017 & 0.6879 & 1.92 & 0.5566 & 0.6075 \\
Image Segmentation & Bagging NBC (10) & Bagging NBC (20) & 0.95 & 0.2418 & 0.8089 & 0.95 & 1.6330 & 0.1778 \\
Image Segmentation & Bagging NBC (10) & Bagging NBC (30) & 0.95 & 0.2418 & 0.8089 & 0.95 & 0.3885 & 0.7174 \\
Image Segmentation & Bagging NBC (20) & Bagging NBC (30) & 0.00 & 0.0000 & 1.0000 & 0.00 & 0.0000 & 1.0000 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:multiple} reports the two multiple-data-set levels. Bagging-30 has the largest average advantage over NBC: 1.50 percentage points at the data-set-averaging level ($z=0.7253$, $p=0.4683$) and 1.51 points at the matched fold-averaging level ($t=1.1336$, $df=6$, $p=0.3002$). Neither result is significant. No comparison among ensemble sizes is significant under either multiple-data-set test; the smallest such p-value is 0.1708 for the matched Bagging-10 versus Bagging-30 comparison.

\begin{table}[H]
\centering
\small
\caption{Multiple-dataset comparisons at the data-set-averaging and matched fold-averaging levels.}
\label{tab:multiple}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llrrrrrr}
\toprule
Baseline & Comparison & Data avg. diff. (\%) & $z$ & $p_z$ & Fold avg. diff. (\%) & $t$ & $p_t$ \\
\midrule
NBC & Bagging NBC (10) & 0.10 & 0.0486 & 0.9613 & 0.11 & 0.0888 & 0.9327 \\
NBC & Bagging NBC (20) & 0.83 & 0.3991 & 0.6899 & 0.82 & 0.9478 & 0.3798 \\
NBC & Bagging NBC (30) & 1.50 & 0.7253 & 0.4683 & 1.51 & 1.1336 & 0.3002 \\
Bagging NBC (10) & Bagging NBC (20) & 0.73 & 0.3496 & 0.7266 & 0.71 & 0.5261 & 0.6266 \\
Bagging NBC (10) & Bagging NBC (30) & 1.40 & 0.6751 & 0.4996 & 1.39 & 1.5987 & 0.1708 \\
Bagging NBC (20) & Bagging NBC (30) & 0.67 & 0.3261 & 0.7444 & 0.69 & 0.4694 & 0.6530 \\
\bottomrule
\end{tabular}
\end{table}

\section{Conclusion}
The current run shows a modest descriptive advantage for larger bagging ensembles: Bagging-30 has the highest mean accuracy on Banknote Authentication and Glass Identification and ties NBC on Image Segmentation. However, every NBC-versus-bagging test has $p>0.05$ at both the single-data-set and multiple-data-set levels. The only significant result is the matched fold comparison between Bagging-10 and Bagging-30 on Glass Identification, and its pooled data-set test is not significant. Consequently, this experiment does not provide consistent statistical evidence that bagging outperforms NBC or that 30 estimators is universally superior. It supports only the narrower conclusion that Bagging-30 has the best observed average performance for these settings and this fixed run.

\end{document}
